---
title: "Using ReacTran"
date: "`r date()`"
---

`ReacTran::tran.1D` returns the reaction-diffusion part of an equation that has a term like
$$\begin{aligned}
    \frac{1}{A} \frac{d}{dx}  A f(x)  ,
\end{aligned}$$
where $f$ is the flux.
This can run into numerical difficulties if $A$ is very small (numerically zero).
In this case, it makes sense to solve, equivalently,
$$\begin{aligned}
    \frac{d}{dx} f(x) + f(x) \frac{d}{dx} \log A 
\end{aligned}$$
I have modified `tran.1D` to do this (in [tran.1D.R](tran.1D.R)),
allowing us to pass in $\log A$ instead of $A$.

Now we need a good way to compute the $\log$ of the cline function 
$$\begin{aligned}
    c(x) = \frac{1}{2}(1+\tanh(-2 x \sqrt(s)/\sigma)) .
\end{aligned}$$
Suppose that $x$ is  large and positive (where the problem is).
We can rewrite
$$\begin{aligned}
1+\tanh(-x) 
&= 1+\frac{e^{-x}-e^{x}}{e^{-x}+e^{x}} \\
&= \frac{2e^{-x}}{e^x+e^{-x}} \\
&= \frac{2e^{-x}}{e^x(1+e^{-2x})} 
&= \frac{2e^{-2x}}{1+e^{-2x}} 
\end{aligned}$$
and so
$$\begin{aligned}
\log(1+\tanh(x) )
&= \log(2) - 2x - \log(1+e^{-2x}) .
\end{aligned}$$

Here's a quick check:
```{r log_tanh}
tanh_cline <- function (x,s,sigma=1,log=FALSE) {
    if (log) {
        - (4*x*sqrt(s)/sigma) - log1p(exp(-4*x*sqrt(s)/sigma))
    } else {
        (1/2)*(1+tanh(-2*x*sqrt(s)/sigma))
    }
}
xx <- seq(-30,50,length.out=200)
# not log-scale
plot(xx,tanh_cline(xx,s=0.1),type='l')
lines(xx,exp(tanh_cline(xx,s=0.1,log=TRUE)),col='red',lty=3,lwd=2)
# log-scale
plot(xx,log(tanh_cline(xx,s=0.1)),type='l',ylim=range(tanh_cline(xx,s=0.1,log=TRUE)))
lines(xx,(tanh_cline(xx,s=0.1,log=TRUE)),col='red',lty=3,lwd=2)
```

Proof of concept
----------------

```{r check_reactran}
library(ReacTran)
orig.tran.1D <- tran.1D
# modified version with log.A option
source("tran.1D.R")

thegrid <- setup.grid.1D(x.up=-30, x.down=30, N = 5000)

Avec <- tanh_cline(thegrid$x.mid,s=1)

# original form
pde.fn <- function (t,y,parms,...) {
    tran <- orig.tran.1D(C=y, A=Avec, D=1/2, dx=thegrid)$dC
    list( tran + (1-Avec) * y )
}

yinit <- rep(1.0,thegrid$N)
times <- seq(0,1,length.out=100)
soln <- ode.1D( y=yinit, times=times, func=pde.fn, nspec=1 )
image(soln,legend=TRUE)

logAvec <- tanh_cline(thegrid$x.mid,log=TRUE,s=1)

pde.fn.log <- function (t,y,parms,...) {
    tran <- tran.1D(C=y, A=logAvec, D=1/2, dx=thegrid, log.A=TRUE)$dC
    list( tran + (1-Avec) * y )
}

soln <- ode.1D( y=yinit, times=times, func=pde.fn.log, nspec=1 )
image(soln,legend=TRUE)
```
